The Unicode Cookbook \linebreak for Linguists

### 2015-09-29 book/chapters/_remove/originaltext.tex
moves book contents into book directory

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005). Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations. For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014). Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

@@

Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/. A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties. The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/. ) Basically, those properties are just a long list of values for each character. For example, code point U+0047 has the following properties (among many others):
\begin{itemize}
	\item Name: LATIN CAPITAL LETTER G 
	\item Alphabetic: YES 
	\item Uppercase: YES 
	\item Script: LATIN 
	\item Extender: NO 
	\item Simple\_Lowercase\_Mapping: 0067 
\end{itemize}

These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes). The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions. For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases). Another use of properties is to define the script of a character. The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems. For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset. The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.

@@

The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used. These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system. In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character. For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish. So, what the user perceives as a single character actually involves a multi-code-point sequence. Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences. The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations. Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}). In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.

@@

The Unicode code space is subdivided into blocks of contiguous code points. For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF. These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems. First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space. For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''. Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there. In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.

The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy. Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character. However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there. For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.

@@

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics. As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen. So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface. Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz. LATIN SMALL LETTER G at U+0067). In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts. All these different A-like characters include the following code points in the Unicode Standard:
\begin{itemize}
	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
	\item <ꓮ> LISU LETTER A, at U+A4EE 
	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
	\item <𐊠> CARIAN LETTER A, at U+102A0 
	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
\end{itemize}

@@

Such homoglyphs are a widespread problem for consistent encoding. Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities. Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type. Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.

@@

Consider for example the characters and following Unicode code points:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
	\item ANGSTROM SIGN (U+212B) 
	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
\end{enumerate}

The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point. In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points. All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance. Nevertheless, they are encoded differently. For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.

@@

It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another. This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original. This might lead to unexpected behaviour for users. Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html. ) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect. For example, Apple's TextEdit does not do any conversion on text entry. However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste. Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)

@@

Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering). For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard. Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261. One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive. Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)

Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD. On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061. But even among linguists this distinction leads to problems. For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984). Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''. The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter). We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts. For instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics. This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph. If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics. This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead. That particular combination of font and software application will thus lead to the desired effect. However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost. In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters. The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized. After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4). This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)

Some other homoglyphs related to encoding IPA in the Unicode Standard are:
\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list. An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019. Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard. An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP . Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online. This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex). This naming is probably best seen as a simple error in the Unicode Standard. Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g. <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7). Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding. However, their usage is discouraged by the Unicode core specification. Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably). Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself. In the context of IPA, characters like the following ligatures should thus \emph{not} be used. Instead a combination of two characters is preferred: 
	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
\end{itemize}

However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
\begin{itemize}
	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
	\item <æ> LATIN SMALL LETTER AE at U+00E6 
\end{itemize}

@@

When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization. Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts. This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters. However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed. In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective. It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.

>>>>

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005). Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations. For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014). Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.

@@

Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/. A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties. The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/. ) Basically, those properties are just a long list of values for each character. For example, code point \uni{0047} has the following properties (among many others):
\begin{itemize}
	\item Name: LATIN CAPITAL LETTER G 
	\item Alphabetic: YES 
	\item Uppercase: YES 
	\item Script: LATIN 
	\item Extender: NO 
	\item Simple\_Lowercase\_Mapping: 0067 
\end{itemize}

These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes). The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions. For example, the corresponding lowercase character to \uni{0047} is \uni{0067} (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases). Another use of properties is to define the script of a character. The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems. For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset. The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.

@@

The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used. These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system. In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character. For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at \uni{006E} followed by COMBINING TILDE at \uni{0303}) combines visually into , a user-percieved character in writing systems like that of Spanish. So, what the user perceives as a single character actually involves a multi-code-point sequence. Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at \uni{00F1}, but this is not the case for all multi-code point character sequences. The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence \uni{006E} \uni{0303} should in all situations be treated identically to the precomposed \uni{00F1}), and a system of \textsc{canonical equivalence} is available for such situations. Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}). In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.

@@

The Unicode code space is subdivided into blocks of contiguous code points. For example, the block called ``Cyrillic'' runs from \uni{0400} till \uni{04FF}. These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems. First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space. For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''. Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there. In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.

The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy. Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character. However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there. For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at \uni{014B} in the block ``Latin Extensions-A''.

@@

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics. As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen. So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface. Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz. LATIN SMALL LETTER G at \uni{0067}). In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts. All these different A-like characters include the following code points in the Unicode Standard:
\begin{itemize}
	\item <A\textgreater{} LATIN CAPITAL LETTER A, at \uni{0041} 
	\item <А> CYRILLIC CAPITAL LETTER A, at \uni{0410} 
	\item <Α> GREEK CAPITAL LETTER ALPHA, at \uni{0391} 
	\item <Ꭺ> CHEROKEE LETTER GO, at \uni{13AA} 
	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at \uni{15C5} 
	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at \uni{1D00} 
	\item <ꓮ> LISU LETTER A, at \uni{A4EE} 
	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at \uni{FF21} 
	\item <𐊠> CARIAN LETTER A, at \uni{102A}0 
	\item <𐌀> OLD ITALIC LETTER A, at \uni{1030}0 
	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, \uni{1D5A}0 
	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at \uni{1D67}0 
\end{itemize}

@@

Such homoglyphs are a widespread problem for consistent encoding. Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely \uni{0076}, \uni{006F}, \uni{0063}, \uni{0065}, \uni{0073}, while the second is a mix of Cyrillic and Greek characters, namely \uni{03BD}, \uni{03BF}, \uni{0041}, \uni{0435}, \uni{0455}.) For computers these two words are completely different entities. Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type. Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.

@@

Consider for example the characters and following Unicode code points:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item LATIN CAPITAL LETTER A WITH RING ABOVE (\uni{00C5}) 
	\item ANGSTROM SIGN (\uni{212B}) 
	\item LATIN CAPITAL LETTER A (\uni{0041}) + COMBINING RING ABOVE (\uni{030A}) 
\end{enumerate}

The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point. In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points. All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance. Nevertheless, they are encoded differently. For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (\uni{212B}), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (\uni{00C5}) would not be found.

@@

It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another. This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original. This might lead to unexpected behaviour for users. Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html. ) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at \uni{0061} and COMBINING GRAVE ACCENT at \uni{0300}, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at \uni{00E0}.\footnote{The behaviour of software applications can be quite erratic in this respect. For example, Apple's TextEdit does not do any conversion on text entry. However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste. Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)

@@

Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering). For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point \uni{0430} for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point \uni{0061}, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard. Another example we commonly encounter is the use of LATIN SMALL LETTER G at \uni{0067}, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at \uni{0261}. One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the \uni{0067} because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive. Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)

Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point \uni{0259} and LATIN SMALL LETTER TURNED E at \uni{01DD}. On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at \uni{0251}, and the open front unrounded vowel , LATIN SMALL LETTER A at \uni{0061}. But even among linguists this distinction leads to problems. For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984). Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''. The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter). We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts. For instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics. This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph. If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics. This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead. That particular combination of font and software application will thus lead to the desired effect. However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost. In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters. The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized. After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4). This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)

Some other homoglyphs related to encoding IPA in the Unicode Standard are:
\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list. An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at \uni{2019}. Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at \uni{02BC}, but unfortunately looks mostly highly similar to \uni{2019}. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard. An example is the \uni{02C1} MODIFIER LETTER REVERSED GLOTTAL STOP vs the \uni{02E4} MODIFIER LETTER SMALL REVERSED GLOTTAL STOP . Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online. This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at \uni{01C3} is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex). This naming is probably best seen as a simple error in the Unicode Standard. Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g. <ʧ> LATIN SMALL LETTER TESH DIGRAPH at \uni{02A7}). Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding. However, their usage is discouraged by the Unicode core specification. Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably). Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself. In the context of IPA, characters like the following ligatures should thus \emph{not} be used. Instead a combination of two characters is preferred: 
	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at \uni{02A3} (use <d> + <z> instead) 
	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at \uni{02A7} (use <t> + <ʃ> instead) 
	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at \uni{02A9} (use <f> + <ŋ> instead) 
\end{itemize}

However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
\begin{itemize}
	\item <ɮ> LATIN SMALL LETTER LEZH at \uni{026E} 
	\item <œ> LATIN SMALL LIGATURE OE at \uni{0153} 
	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at \uni{0276} 
	\item <æ> LATIN SMALL LETTER AE at \uni{00E6} 
\end{itemize}

@@

When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization. Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at \uni{00E1}) will be split into its parts. This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters. However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed. In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective. It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
